---
title: "STA302-A1"
output:
  word_document: default
  pdf_document: default
editor_options: 
  chunk_output_type: console
---

## Ishaan Nagi
##1002452525

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

SETUP
```{r}
setwd("~/Google Drive/Uni/Winter 2018/STA302/A1")
Dat=read.table("GPA.txt", header=F) 
# change column names 
names(Dat)<-c("GPA","ACT")
# print the six first observations head(Dat)
Dat <- Dat[, c("ACT", "GPA")]
n = nrow(Dat)
```


(a)
```{r}
gpa <- Dat[,2]
act <- Dat[,1]
mu_hat_act <- mean(act)
mu_hat_gpa <- mean(gpa)
var_gpa <- var(gpa)
var_act <- var(act)
```
mu_hat_act = 24.725
mu_hat_gpa =  3.07405
var_act = 19.99937
var_gpa = 0.4151719



(b) 
```{r}
corr_ACT_GPA <- cor(gpa, act)
```
corr_ACT_GPA = 0.2694818;
Weak positive correlation.



(c)
```{r}
x <- act
y <- gpa
fit.Gpa <- lm(GPA ~ ACT, data=Dat)
```


(d) b0(intercept) = 2.11404929, b1(slope) =  0.03882713.
[from coef(fit.Gpa) output]



(e) For every unit change in the ACT score there is an change in the GPA score by 0.3882713.



(f)
```{r}
b0 = 2.11404929
b1 = 0.03882713 
RSS <-sum(resid(fit.Gpa)^2)
s_square = RSS / (n-2)
s = (s_square)^(0.5)
```
s = 0.623125 is the SD around regression line.



(g)

```{r}
# H_0: B1 = 0
# H_A: B1 != 0

y_mean <- mu_hat_gpa
x_mean <- mu_hat_act
std_error_B1 <- ((s_square)/(sum((x - x_mean)^2)))^(1/2)
t_stat <- (b1) / (std_error_B1)
```
t-stat =  3.040
t-value = 1.980	
Since t-value < t-stat, we reject the H_0.
Then b1 is not 0 and there is a linear relationship between ACT scores and GPA.



(h)
```{r}
coef(fit.Gpa) #b0 and b1
plot(y~x, data=Dat, col="red", pch=19, xlab="ACT", ylab="GPA", xlim=c(0,40))
abline(coef(fit.Gpa), col="orange")
abline(v=x_mean, h=y_mean, lty=2)
title('Scatter PLot and the Line of the best fit for ACT and GPA data')
```

(i)
```{r}
anova(fit.Gpa)
```

(j)

```{r}
mse <- RSS/n-2
# [from table]
```
mse = RSS/n-2 = 0.3883


(k)

```{r}
# H_0: B1 = 0
# H_A: B1 != 0
```

f-value = 0.002917
f(.05, 1, 118) = 3.9201	#f-stat
Since f-value < f-stat, we reject the H_0.
Then b1 is not 0 and there is a linear relationship between ACT scores and GPA.


(l)
The results in part k and g are the same.


(m)
```{r}
r.squared <- 0.0726
#[From str(summary(fit.Gpa)) output]
```
Only 7.26% of the variation in GPA explained by the regression model


(n)
```{r}
gpa <- data.frame(ACT=28)
CI_gpa_28 <- predict(fit.Gpa, gpa, interval="confidence", level = 0.95, se.fit = TRUE)
```
CI_GPA_28 -> [3.061384, 3.341033]
We are 95% confident that the mean GPA of students whose ACT score was 28 will lie within the above bounds.



(o)
```{r}
PI_gpa_20 <- predict(fit.Gpa, gpa, interval="prediction", level = 0.95, se.fit = TRUE)
```
PI_GPA_20 -> [1.645753 4.13543]
We can Predict that a GPA of a student with ACT score 20 will have in the range given above. [95% of the time]