---
title: "STA302A2"
output:
  pdf_document: default
  word_document: default
editor_options:
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

##ISHAAN NAGI 
##1002452525

SETUP
```{r}
setwd("~/Google Drive/Uni/Winter 2018/STA302/A2")
dat <- read.table("Census.txt", sep = "" ,header=T)
LIFE <- dat$LIFE
MALE <- dat$MALE
BIRTH <- dat$BIRTH
DIVO <- dat$DIVO
BEDS <- dat$BEDS
EDUC <- dat$EDUC
INCO <- dat$INCO
y <- LIFE
```

1)
```{r}
x = as.matrix(dat)
x <-x[1:51, 2:7]
x<- cbind(1, x)
mode(x) = 'numeric'
n <- 51
p <-6
X<-x
Xt <- t(X)
XtX <- Xt %*% X
XtXinv <-solve(XtX)
Xty <- Xt %*% y
bhat <- XtXinv %*% Xty
```

  1a)
```{r}
y_hat <- X %*% bhat
y_hat
```

```{r}

e_hat = y - y_hat
e_hat
```

```{r}
bhat
```

Equation:
LIFE = 70.5577812705 + 0.1261018758 (MALE) - 0.5160557876 (BIRTH) - 0.1965375074 (DIVO) - 0.0033392036 (BEDS) + 0.2368222541 (EDUC) - 0.0003612011 (INCO) + e_hat
  
  
  
  1b)
  - MALE
```{r}
  bhat[2] #b_1
```

b_1 (MALE) corresponds to the Expected change (+ 0.1261019) in the Averafe Lifespan with 1 unit increase in the proportion of Males to Female.
  
 - BIRTH
```{r}
bhat[3] #b_2
```

b_2 (BIRTH) corresponds to the Expected change (-0.5160558) in the Averafe Lifespan with a unit increase in the birth rate per 1,000 people.
  
  
  1c)
  
```{r}
RSS <- (t(e_hat) %*% e_hat)
sigma_sq_hat = RSS / (n)
sigma_sq_hat
```



```{r}
s_sq = RSS / (n-p-1)
s_sq
```

  1d)
  
```{r}
s <-c(s_sq^(1/2))
se_bs = diag(s * (XtXinv)^(1/2))
se_bs
#Corresponding to B_0, B_1 ... B_6
```


  1e)

```{r}
y_bar = c(sum(y_hat)/ n)
SST <- sum((LIFE - y_bar)^(2))
R_sq <- 1 - (RSS/SST)
R_sq
```

Explains the proportion of variation in the Average lifespan explained by the regression, which is 46.849%.


2)
  2a)
```{r}
MLR <- lm(formula = LIFE~ MALE + BIRTH + DIVO + BEDS + EDUC + INCO, data=dat) 
summary(MLR)
```


  2b)
  $H_0: B_0 = B_1 = B_2 = B_3 = B_4 = B_5 = B_6 = 0$
  $H_A: B_i \neq 0$ [any i in range 0 to 6]
```{r}
SSReg = SST-RSS
F_val = (SSReg/(p)/(RSS/(n-p-1))) 
F_val
F_crt = qf(.95, df1=p, df2=n-p-1)
F_crt
```

F_val > F_crit, hence we fail to accept $H_0$. Then atleast one of the B_i's [i = 1, ... 6] is not Zero.
Our Model is significant.

  2c)
  
#MALE
  $H_0: B_1 = 0$
  $H_A: B_1 \neq 0$
```{r}
t_value1 =2.670
t_crit = qt(c(.005, .995), df=n-p-1)
#t_crit = [-2.692278, 2.692278]
```

Since |t_crit| > |t_value1|, we accept $H_0$. Then $B_1 \neq 0$ and we can remove the predictor (MALE) from the model.

#BIRTH
  $H_0: B_2 = 0$
  $H_A: B_2 != 0$
```{r}
t_value2 = 4.400
t_crit = qt(c(.005, .995), df=n-p-1)
#t_crit = [-2.692278, 2.692278]
```

Since |t_value2| > |t_crit|, we fail to accept $H_0$. Then $B_2 \neq 0$ then we can't remove the predictor (BIRTH) variable from the model. 

#DIVO
  $H_0: B_3 = 0$
  $H_A: B_3 \neq 0$
```{r}
t_value3 =2.658
t_crit = qt(c(.005, .995), df=n-p-1)
#t_crit = [-2.692278, 2.692278]
```

Since |t_crit| > |t_value3|, we accept $H_0$. Then $B_3 \neq 0$ and we can remove the predictor (DIVO) from the model.

#BEDS
  $H_0: B4 = 0$
  $H_A: B4 \neq 0$
```{r}
t_value4 =3.409
t_crit = qt(c(.005, .995), df=n-p-1)
#t_crit = [-2.692278, 2.692278]
```

Since |t_value4| > |t_crit|, we fail to accept $H_0$. Then $B_4 \neq 0$ then we can't remove the predictor (BEDS) variable from the model. 


#EDUC
  $H_0: B5 = 0$
  $H_A: B5 \neq 0$
```{r}
t_value5 = 2.133
t_crit = qt(c(.005, .995), df=n-p-1)
#t_crit = [-2.692278, 2.692278]
```

Since |t_crit| > |t_value5|, we accept $H_0$. Then $B_3 \neq 0$ and we can remove the predictor (EDUC) from the model.

#INCO
  $H_0: B6 = 0$
  $H_A: B6 \neq 0$
```{r}
t_value6 =0.786
t_crit = qt(c(.005, .995), df=n-p-1)
#t_crit = [-2.692278, 2.692278]
```

Since |t_crit| > |t_value6|, we accept $H_0$. Then $B_6 \neq 0$ and we can remove the predictor (INCO) from the model.

Yes, the results indicate that variables  MALE (x1), DIVO (x3), EDUC (x5) and INCO x(6) should be removed.


  2d)
  
```{r}
MLR_reduced <- lm(formula = LIFE~ BIRTH + BEDS, data=dat) 
summary(MLR_reduced)
```

LIFE = 79.1473186 - 0.3281679 (BIRTH) - 0.0027415 (BEDS)


  2e)
  $H_0: B_1 = B_6 = 0$
  $H_A: B_1, B_6$ both not zero.
  
```{r}
MLR_red_MALE_INCO <- lm(formula = LIFE~ BIRTH + BEDS + DIVO + EDUC, data=dat) 
anova(MLR, MLR_red_MALE_INCO)
F_val = 3.5642
F_Crit =  qf(.99, df1=2, df2=n-p-1)
#F_Crit = 5.122628
```

|F_val| > |F_crit| we fail to reject $H_0$, hence the predictors MALE and INCO can be removed from the model. 

  2f)
```{r}
MLR_MALE <- lm(LIFE ~ MALE, data=dat) 
anova(MLR, MLR_MALE)
F_val = 7.0963
```

As per ANOVA Output, our F test Statistic is 7.0963.

  2g)
  $H_0: B_i = 0$ [i = 1, 2]
  $H_A:$ $B_1, B_2$ , both not zero.
  
```{r}
MLR_B0 = lm(LIFE ~ 1, data=dat) 
MLR_MALE_BIRTH <- lm(LIFE ~ MALE + BIRTH, data=dat) 
anova(MLR_B0, MLR_MALE_BIRTH)
F_val = 8.14
```

We fail to accept the $H_0$, we infer that (MALE) x1 and(BIRTH) x2 are significant/useful in predicting the response.


  2h)
```{r}
MLR_B0 = lm(LIFE ~ 1, data=dat) 
MLR_B0B3 = lm(LIFE ~ DIVO)
MLR_B0B2B3 = lm(LIFE ~ BIRTH + DIVO)
MLR_MALE_BIRTH_DIVO <- lm(LIFE~ MALE + BIRTH + DIVO, data=dat) 
MLR_BIRTH_DIVO <- lm(LIFE ~ BIRTH + DIVO, data=dat) 


```

```{r}
SSRegB3B0 <- anova(MLR_B0)["Residuals", "Sum Sq"] - anova(MLR_B0B3)["Residuals", "Sum Sq"]
SSRegB3B0
```


```{r}
SSRegB2_B0B3 <- anova(MLR_B0B3)["Residuals", "Sum Sq"] - anova(MLR_B0B2B3)["Residuals", "Sum Sq"]
SSRegB2_B0B3
```


```{r}
RSSB0B1B2B3 <- anova(MLR_MALE_BIRTH_DIVO)["Residuals", "Sum Sq"]

RSS_B0B2B3 <- anova(MLR_BIRTH_DIVO)["Residuals", "Sum Sq"] 

SSRegB1_B0B2B3 <- RSS_B0B2B3 - RSSB0B1B2B3
SSRegB1_B0B2B3

```


```{r}
SSRegB1B2B3_B0 =  SSRegB3B0 + SSRegB2_B0B3 + SSRegB1_B0B2B3
SSRegB1B2B3_B0
```

  2i)
      - 2i) (1)
    
```{r}
#MLR_B0
#MLR_MALE
MLR_BIRTH <- lm(LIFE ~ BIRTH, data=dat) 
MLR_DIVO <- lm(LIFE ~ DIVO, data=dat) 
MLR_INCO <- lm(LIFE ~ INCO, data=dat) 
#MLR_MALE_BIRTH
MLR_MALE_DIVO <- lm(LIFE~ MALE + DIVO, data=dat) 
MLR_MALE_INCO <- lm(LIFE~ MALE + INCO, data=dat) 
#MLR_BIRTH_DIVO
MLR_BIRTH_INCO <- lm(LIFE ~ BIRTH + INCO, data=dat) 
MLR_DIVO_INCO <- lm(LIFE ~ DIVO + INCO, data=dat) 
MLR_MALE_BIRTH_DIVO
MLR_MALE_BIRTH_INCO <- lm(LIFE~ MALE + BIRTH + INCO, data=dat) 
MLR_BIRTH_DIVO_INCO <- lm(LIFE ~ BIRTH + DIVO + INCO, data=dat) 
MLR_MALE_DIVO_INCO <- lm(LIFE ~ MALE + DIVO + INCO, data=dat) 
MLR_mbdi <- lm(LIFE ~ MALE + BIRTH + DIVO + INCO, data=dat) 

#AIC
aic_b0 <- AIC(MLR_B0) 
aic_b0
aic_m <- AIC(MLR_MALE) 
aic_m
aic_d <- AIC(MLR_DIVO)
aic_d
aic_i <- AIC(MLR_INCO)
aic_i
aic_b <- AIC(MLR_BIRTH)
aic_b
aic_mb <- AIC(MLR_MALE_BIRTH)
aic_mb
aic_md <- AIC(MLR_MALE_DIVO)
aic_md
aic_mi <- AIC(MLR_MALE_INCO)
aic_mi
aic_bd <- AIC(MLR_BIRTH_DIVO)
aic_bd
aic_bi <- AIC(MLR_BIRTH_INCO)
aic_bi
aic_di <- AIC(MLR_DIVO_INCO)
aic_di
aic_mbd <- AIC(MLR_MALE_BIRTH_DIVO)
aic_mbd
aic_mbi <- AIC(MLR_MALE_BIRTH_INCO)
aic_mbi
aic_bdi <- AIC(MLR_BIRTH_DIVO_INCO)
aic_bdi
aic_mdi <- AIC(MLR_MALE_DIVO_INCO)
aic_mdi
aic_mbdi <- AIC(MLR_mbdi)
aic_mbdi
```

Lowest AIC is with model LIFE ~ BIRTH + MALE + DIVO

      - 2i) (2)
```{r}
null<- MLR_B0
full<- MLR_mbdi
forwdAIC=step(null, scope=list(lower=null, upper=full), direction="forward")
forwdAIC
```

Best Model according to forwardAIC is LIFE ~ BIRTH + MALE + DIVO

      - 2i) (3)
```{r}
backAIC=step(full, direction="backward", data=dat)
backAIC
```

Best Model according to backwardAIC is LIFE ~ BIRTH + MALE + DIVO


3)

  3a)
  
```{r}
std_res <- rstandard(MLR)
plot(MLR$fitted.values, std_res, ylab="Standardized Residuals", xlab="Fitted Values", main="Standardized Residuals")
abline(0, 0)
abline(2, 0, col=c("red"))
abline(-2, 0, col=c("red"))
```

Yes, there are 3 points with residuals greater than |2| and an outlier that may be influential with the lowest fitted value (~67).

  3b)
```{r}
lev = hat(model.matrix(MLR))
plot(lev)
abline(0.5, 0, col=c("red"))
dat[lev >0.5,]
```

3 points have lev > 0.5. Indices 1, 8 and 34. Corresponding to States: AK, DC and NV.

  3c)
```{r}
cook = cooks.distance(MLR)
plot(cook)
dat[cook > 1,]
```

No, not all Observations are the same. We now see data corresponding to UT instead of NV.


  3d)
```{r}
plot(BEDS, std_res, ylab="Standardized Residuals", 
     xlab="BEDS", main="Standardized Residuals")
abline(0, 0)
abline(2, 0, col=c("red"))
abline(-2, 0, col=c("red"))
points(BEDS[8], std_res[8] , col='red', pch = 19)
```

There are 3 points that are outliers based on std_res (>2) and the outlier red point corresponding to DC (RED)


  3e)

```{r}
dat2 <- subset(dat, STATE != 'DC')
LIFE2 <- dat2$LIFE
MALE2 <- dat2$MALE
BIRTH2 <- dat2$BIRTH
DIVO2 <- dat2$DIVO
BEDS2 <- dat2$BEDS
EDUC2 <- dat2$EDUC
INCO2 <- dat2$INCO
MLR_NoDC <- lm(formula = LIFE2 ~ MALE2 + BIRTH2 + DIVO2 + BEDS2 + EDUC2 + INCO2, data=dat2)
summary(MLR_NoDC)
std_res_NoDC <- rstandard(MLR_NoDC)
plot(MLR_NoDC$fitted.values, std_res_NoDC, ylab="Standardized Residuals", xlab="Fitted Values", main="Standardized Residuals")
abline(0, 0)
abline(2, 0, col=c("red"))
abline(-2, 0, col=c("red"))
```


Yes there is a noticeable change in the Standardised residuals. 
The fitted line has changed.
NEW:
  - LIFE = 69.8207566 + 0.0922054 (MALE2) - 0.4261398 (BIRTH2) - 0.1377833 (DIVO2) - 0.0011637 (BEDS2) + 0.3157699 (EDUC2) - 0.0004698 (INCO2)
OLD:
- LIFE = 70.5577813 + 0.1261019 (MALE2) - 0.5160558 (BIRTH2) - 0.1965375 (DIVO2) - 0.0033392 (BEDS2) + 0.2368223 (EDUC2) - 0.0003612 (INCO2)

Also, $R^2$s changed:
New: Multiple R-squared:  0.3679
Old: Multiple R-squared:  0.4685

  f)
  
```{r}
#MALE
plot(MALE2, std_res_NoDC, ylab="Standardized Residuals", xlab="MALE", main="Standardized Residuals")
abline(0, 0)
abline(2, 0, col=c("red"))
abline(-2, 0, col=c("red"))
points(MALE[44], std_res_NoDC[44] , col='red', pch = 19)

#BEDS
plot(BIRTH2, std_res_NoDC, ylab="Standardized Residuals", xlab="BIRTH", main="Standardized Residuals")
abline(0, 0)
abline(2, 0, col=c("red"))
abline(-2, 0, col=c("red"))
points(BIRTH2[44], std_res_NoDC[44] , col='red', pch = 19)

#DIVO
plot(DIVO2, std_res_NoDC, ylab="Standardized Residuals", xlab="DIVO", main="Standardized Residuals")
abline(0, 0)
abline(2, 0, col=c("red"))
abline(-2, 0, col=c("red"))
points(DIVO[44], std_res_NoDC[44] , col='red', pch = 19)

#BEDS
plot(BEDS2, std_res_NoDC, ylab="Standardized Residuals", xlab="BEDS", main="Standardized Residuals")
abline(0, 0)
abline(2, 0, col=c("red"))
abline(-2, 0, col=c("red"))
points(BEDS2[44], std_res_NoDC[44] , col='red', pch = 19)

#EDUC
plot(EDUC2, std_res_NoDC, ylab="Standardized Residuals", xlab="EDUC", main="Standardized Residuals")
abline(0, 0)
abline(2, 0, col=c("red"))
abline(-2, 0, col=c("red"))
points(EDUC2[44], std_res_NoDC[44] , col='red', pch = 19)

#INCO
plot(INCO2, std_res_NoDC, ylab="Standardized Residuals", xlab="INCO", main="Standardized Residuals")
abline(0, 0)
abline(2, 0, col=c("red"))
abline(-2, 0, col=c("red"))
points(INCO2[44], std_res_NoDC[44] , col='red', pch = 19)
```
ALL Data corresponding to this STATE has very standardized residual values.
Difference between observations and expected values is large.



  3g)
```{r}
dat3 <- subset(dat2, STATE != 'UT')
LIFE3 <- dat3$LIFE
MALE3 <- dat3$MALE
BIRTH3 <- dat3$BIRTH
DIVO3 <- dat3$DIVO
BEDS3 <- dat3$BEDS
EDUC3 <- dat3$EDUC
INCO3 <- dat3$INCO
MLR_NoDC_NoUT <- lm(formula = LIFE3 ~ MALE3+ BIRTH3 + DIVO3 + BEDS3 + EDUC3 + INCO3, data=dat3)
```


$R^2$ is higher without UT data. Therefore this model explains the proportion of variability explained by the regression, better.
Predictor have coefficients chnaged:
```{r}
MLR_NoDC_NoUT$coefficients
```

Vs
```{r}
MLR$coefficients
```


